A tool set for random number generation on GPUs in R

### 2021-12-28 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
paper 1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3072 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is above 160 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 512 * 128)
R> system.time(clrng::rnorm(c(10000, 10000), streams = streams,
+    Nglobal = c(512, 128), type = "double"))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.052   0.000   0.051
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 256 * 64)
R> month_gpu <- vclMatrix(month, type = "integer")
R> system.time(result_month <- clrng::fisher.sim(month_gpu, 1000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.321   0.003   0.324
\end{CodeOutput}
\begin{CodeInput}
R> result_month$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -47955
\end{CodeOutput}
\begin{CodeInput}
R> result_month$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 1015808
\end{CodeOutput}
\begin{CodeInput}
R> result_month$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 409435
\end{CodeOutput}
\begin{CodeInput}
R> result_month$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.403
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> week_GPU <- gpuR::vclMatrix(week, type = "integer")
R> system.time(result_week <- clrng::fisher.sim(week_GPU, 10000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   1.934   0.021   1.953
\end{CodeOutput}
\begin{CodeInput}
R> result_week$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -54990
\end{CodeOutput}
\begin{CodeInput}
R> result_week$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 10010624
\end{CodeOutput}
\begin{CodeInput}
R> result_week$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 1351
\end{CodeOutput}
\begin{CodeInput}
R> result_week$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.000135
\end{CodeOutput} 
\end{CodeChunk}

@@

% $U=L*D^{1/2}*Z$, where 
% \begin{itemize}
% \itemsep0em 
% \item $L$ is the lower unit triangular matrix in Cholesky decomposition of $\Sigma = L*D*L^\top$,
% \item $D$ is a diagonal matrix.
% \item $Z=(Z_1, Z_2, \dots, Z_n) \sim \text{MVN}(0,I_n)$, where $I_n$ is a $n \times n$ identity matrix.
% \end{itemize}
The matrix decomposition method is straightforward to implement. However, when the number of locations $n$ is large, the cost of Cholesky decomposition of the covariance matrix is $\mathcal{O}(n^3)$,  and the cost of matrix-vector multiplication $L*Z$ to generate the random field $U$ is $\mathcal{O}(n^2)$ \citep{LiuandLi2019}.  Our other \proglang{R} package \pkg{gpuBatchMatrix} is able to compute batches of Mat\'ern covariance matrices in parallel, and do Cholesky decomposition and matrix-matrix multiplication in batches in parallel on GPU. The following shows an example, in which we simulate on GPU eight Gaussian random fields of Mat\'ern covariance at one time with four sets of parameters, by taking advantage of the GPU-capabilities offered by \pkg{clrng} and \pkg{gpuBatchMatrix} together.
%\pkg{clrng} package not only does the matrices decomposition (Cholesky decomposition and matrix-vector multiplication) in parallel, but also computes the covariance matrices in parallel on GPU.

@@

Step 1, set up a $51 \times 80$ raster using the 'swissRain' data from \pkg{geostatsp}.
\begin{CodeChunk}
\begin{CodeInput}
R> NlocalCache = 1000
R> Nglobal = c(128, 64, 2)
R> Nlocal = c(4, 2, 2)
R> theType = "double"
\end{CodeInput} 
\end{CodeChunk}

@@

Step 2, create 4 parameter sets as a small example.

@@

Step 4, perform Cholesky decomposition of \code{maternCov}. The first argument in \fct{cholBatch} specifies the object to take Cholesky decomposition. Computed unit lower triangular matrices $L_i$'s are stacked by row and stored in \code{maternCov}. The diagonal values of each $D_i$ are returned and stored in each row of \code{diagMat}. For example, if each batch $\Sigma_i$ is of size $n \times n$, then each batch $D_i$ is $1 \times n$.
\begin{CodeChunk}
\begin{CodeInput}
R> diagMat = gpuBatchMatrix::cholBatch(maternCov, Nglobal = c(128,
+    8), Nlocal = c(32, 8))
\end{CodeInput} 
\end{CodeChunk}

\begin{gather}
\begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
\end{bmatrix}
 \rightarrow
 \begin{bmatrix}
  L_{1} \\ L_{2} \\L_{3} \\ \vdots
  \end{bmatrix} \text{and}
  \begin{bmatrix}
  D_{1} \\ D_{2} \\D_{3} \\ \vdots
  \end{bmatrix}
\end{gather}
Step 5, create some streams on GPU to generate 2 standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{clrng::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
\begin{CodeChunk}
\begin{CodeInput}
R> streamsGpu <- createStreamsGpu(n = 128 * 64)
\end{CodeInput} 
\end{CodeChunk}

@@

\begin{gather}
 \begin{bmatrix}  L_{1} \\ L_{2} \\L_{3} \\ \vdots
 \end{bmatrix} 
 *
  \begin{bmatrix}
   D_{1} \\ D_{2} \\D_{3} \\ \vdots
   \end{bmatrix} 
   *
   \begin{bmatrix}
   Z_{11} & Z_{12}
   \end{bmatrix}
  =
 \begin{bmatrix}
   L_1D_1Z_{11} & L_1D_1Z_{12} \\
   L_2D_2Z_{11} & L_2D_2Z_{12} \\
   L_3D_3Z_{11} & L_3D_3Z_{12} \\
   \vdots  &   \vdots
  \end{bmatrix}
\end{gather}
Step 7, Finally, plot the 10 realizations.
\begin{CodeChunk}
\begin{CodeInput}
R> simRaster = raster::brick(myRaster, nl = ncol(simMat) * nrow(params))
R> values(simRaster) = as.vector(as.matrix(simMat))
\end{CodeInput} 
\end{CodeChunk}

>>>>

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is above 160 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 512 * 128)
R> system.time(clrng::rnorm(c(10000, 10000), streams = streams,
+    Nglobal = c(512, 128), type = "double"))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.051   0.001   0.052
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> streams <- createStreamsGpu(n = 256 * 64)
R> month_gpu <- vclMatrix(month, type = "integer")
R> system.time(result_month <- clrng::fisher.sim(month_gpu, 1000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   0.315   0.008   0.322
\end{CodeOutput}
\begin{CodeInput}
R> result_month$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -47955
\end{CodeOutput}
\begin{CodeInput}
R> result_month$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 1015808
\end{CodeOutput}
\begin{CodeInput}
R> result_month$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 409435
\end{CodeOutput}
\begin{CodeInput}
R> result_month$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.403
\end{CodeOutput} 
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> week_GPU <- gpuR::vclMatrix(week, type = "integer")
R> system.time(result_week <- clrng::fisher.sim(week_GPU, 10000000,
+    streams = streams, type = "double", returnStatistics = TRUE,
+    Nglobal = c(256, 64)))
\end{CodeInput}
\begin{CodeOutput}
##    user  system elapsed 
##   1.927   0.019   1.945
\end{CodeOutput}
\begin{CodeInput}
R> result_week$threshold
\end{CodeInput}
\begin{CodeOutput}
## [1] -54990
\end{CodeOutput}
\begin{CodeInput}
R> result_week$simNum
\end{CodeInput}
\begin{CodeOutput}
## [1] 10010624
\end{CodeOutput}
\begin{CodeInput}
R> result_week$counts
\end{CodeInput}
\begin{CodeOutput}
## [1] 1351
\end{CodeOutput}
\begin{CodeInput}
R> result_week$p.value
\end{CodeInput}
\begin{CodeOutput}
## [1] 0.000135
\end{CodeOutput} 
\end{CodeChunk}

@@

% $U=L*D^{1/2}*Z$, where 
% \begin{itemize}
% \itemsep0em 
% \item $L$ is the lower unit triangular matrix in Cholesky decomposition of $\Sigma = L*D*L^\top$,
% \item $D$ is a diagonal matrix.
% \item $Z=(Z_1, Z_2, \dots, Z_n) \sim \text{MVN}(0,I_n)$, where $I_n$ is a $n \times n$ identity matrix.
% \end{itemize}
The downside of this method is that it's computationally expensive and is thus very slow. When the number of locations $n$ is large, the time cost of Cholesky decomposition of the covariance matrix is $\mathcal{O}(n^3)$,  and is $\mathcal{O}(n^2)$ for the matrix-vector multiplication $L*Z$ \citep{LiuandLi2019}.  Our other \proglang{R} package \pkg{gpuBatchMatrix} is able to address this issue since it computes batches of Mat\'ern covariance matrices in parallel, and does Cholesky decomposition and matrix-matrix multiplication in batches in parallel on GPU. The following shows an example, in which we simulate on GPU eight Gaussian random fields of Mat\'ern covariance at one time with four sets of parameters, by taking advantage of the GPU-capabilities offered by \pkg{clrng} and \pkg{gpuBatchMatrix} together.
%\pkg{clrng} package not only does the matrices decomposition (Cholesky decomposition and matrix-vector multiplication) in parallel, but also computes the covariance matrices in parallel on GPU.

@@

Step 1, set up a $51 \times 80$ raster using the `swissBorder' coordinates from \pkg{geostatsp}.
\begin{CodeChunk}
\begin{CodeInput}
R> NlocalCache = 1000
R> Nglobal = c(128, 64, 2)
R> Nlocal = c(4, 2, 2)
R> theType = "double"
\end{CodeInput} 
\end{CodeChunk}

@@

Step 2, create four parameter sets as a small example.

@@

Step 4, perform Cholesky decomposition on \code{maternCov}. The first argument in \fct{cholBatch} specifies the object to take Cholesky decomposition. Computed unit lower triangular matrices $L_i$'s are stacked by row in \code{maternCov}. The diagonal values of each $D_i$ are returned and stored in each row of \code{diagMat}. So if each batch $\Sigma_i$ is of size $n \times n$, then each batch $D_i$ is $1 \times n$.
\begin{CodeChunk}
\begin{CodeInput}
R> diagMat = gpuBatchMatrix::cholBatch(maternCov, Nglobal = c(128,
+    8), Nlocal = c(32, 8))
\end{CodeInput} 
\end{CodeChunk}

\begin{gather}
\begin{bmatrix} \Sigma_{1} \\ \Sigma_2 \\ \Sigma_3 \\ \vdots
\end{bmatrix}
 \rightarrow
 \begin{bmatrix}
  L_{1} \\ L_{2} \\L_{3} \\ \vdots
  \end{bmatrix} \text{and}
  \begin{bmatrix}
  D_{1} \\ D_{2} \\D_{3} \\ \vdots
  \end{bmatrix}
\end{gather}
Step 5, create some streams on GPU to generate two standard Gaussian random vectors \code{zmatGpu}$=(Z_1, Z_2)$ using \fct{clrng::rnorm}, in which \code{c(nrow(materCov),2)} specifies the number of rows and columns of \code{zmatGpu}.
\begin{CodeChunk}
\begin{CodeInput}
R> streamsGpu <- createStreamsGpu(n = 128 * 64)
\end{CodeInput} 
\end{CodeChunk}

@@

\begin{gather}
 \begin{bmatrix}  L_{1} \\ L_{2} \\L_{3} \\ \vdots
 \end{bmatrix} 
 *
  \begin{bmatrix}
   D_{1} \\ D_{2} \\D_{3} \\ \vdots
   \end{bmatrix} 
   *
   \begin{bmatrix}
   Z_{11} & Z_{12}
   \end{bmatrix}
  =
 \begin{bmatrix}
   L_1D_1Z_{11} & L_1D_1Z_{12} \\
   L_2D_2Z_{11} & L_2D_2Z_{12} \\
   L_3D_3Z_{11} & L_3D_3Z_{12} \\
   \vdots  &   \vdots
  \end{bmatrix}
\end{gather}
Step 7, Finally, plot the ten realizations.
\begin{CodeChunk}
\begin{CodeInput}
R> simRaster = raster::brick(myRaster, nl = ncol(simMat) * nrow(params))
R> values(simRaster) = as.vector(as.matrix(simMat))
\end{CodeInput} 
\end{CodeChunk}
